Optimizing and Incrementalizing Collection Queries by AST Transformations

### 2017-06-02 aosd13/pepm13/caseStudyImpl.tex
Chap. 5: Reformat code/revise a bit

With these declarations in scope, the Scala compiler rewrites \code{str1 + str2} to
\code{expToStringOps(str1).+(str2)}, which evaluates to \code{StringConcat(str1, str2)} as desired.
Note that the implicit conversion function \code{expToStringOps} is not applicable to \code{Exp[Boolean]} because it explicitly specifies the receiver of the \code{+}-call to have type \code{Exp[String]}.
In other words, expressions like \code{str1 + str2} are now \emph{lifted} on the level of expression trees in a type-safe way. For brevity, we refer to the defined operator as \code{Exp[String].+}.

@@

The compiler will now rewrite \code{str1 + "foo"} to \code{expToStringOps(str1) + pure("foo")}.
%Since \code{Exp[String].+} requires an argument of type \code{Exp[String]} but \code{"foo"} has type \code{String},
%the compiler wraps \code{"foo"} with a call to \code{pure}. Hence
Similarly, \code{author. firstName + " " + author.lastName} is rewritten to \code{(expToStringOps(author.firstName) + pure(" ")) + author.lastName}.
Note how the implicit conversions cooperate successfully to lift the expression.

@@

\code{Author} is not part of {\LoS} or the standard Scala library but an application-specific class, hence it is not possible
to pre-define the necessary lifting code as part of {\LoS}. Instead, the application programmer needs to provide this code to connect {\LoS} to his application. To support the application programmer with this tedious task, we provide a code generator which discovers the interface of a class through reflection on its compiled version and generates the boilerplate code such as the one above for \code{Author}. It also generates 
the application-specific expression tree types such as \code{AuthorFirstName} as shown in Sec.~\ref{subsec:repexptrees}.
In general, query writers need to generate and import the boilerplate lifting code for all application-specific types they want to use in a {\LoS} query.

If desired, we can exclude some methods to \emph{restrict} the language supported in our deeply embedded programs. For instance \LoS{} requires the user to write side-effect-free queries, hence we do not lift methods which perform side effects.

@@

An expression like \code{"foo" + "bar" + str1} is converted to \code{toStringOps("foo" + "bar") + str1}; hence, part of the expression is evaluated before being reified. This is harmless here since we want \code{"foo" + "bar"} to be evaluated at compile-time, that is constant-folded, but in other cases it is preferable to prevent the constant folding. We will see later examples where queries on collections are evaluated before reification, defeating the purpose of our framework, and how we work around those.

@@

%\begin{lstlisting}
%def toFOAS[S, T](funBody: Exp[S] => Exp[T]) = 
%\end{lstlisting}
In our example, \code{Fun(str => str + "!")} produces \code{FOASFun(StringConcat(TypedVar[String](1), Const("!")), TypedVar[String](1))}%
%: since an unbound variable appears here, this can be easily recognized as an open term
.

>>>>

With these declarations in scope, the Scala compiler rewrites \code{str1 + str2} to
\code{expToStringOps(str1).+(str2)}, which evaluates to \code{StringConcat(str1, str2)} as desired.
The implicit conversion function \code{expToStringOps} is not applicable to \code{Exp[Boolean]} because it explicitly specifies the receiver of the \code{+}-call to have type \code{Exp[String]}.
In other words, expressions like \code{str1 + str2} are now \emph{lifted} on the level of expression trees in a type-safe way. For brevity, we refer to the defined operator as \code{Exp[String].+}.

@@

The compiler will now rewrite \code{str1 + "foo"} to \code{expToStringOps(str1) + pure("foo")}, and
\code{str1 + "foo" + str2} to \code{expToStringOps(str1) + pure("foo") + str2}.
Different implicit conversions cooperate successfully to lift the expression.
%Since \code{Exp[String].+} requires an argument of type \code{Exp[String]} but \code{"foo"} has type \code{String},
%the compiler wraps \code{"foo"} with a call to \code{pure}. Hence

@@

Implicit conversions for user-defined types cooperate with other ones; for instance, \code{author.firstName + " " + author.lastName} is rewritten to
\begin{lstlisting}
(expToStringOps(expToAuthorOps(author).firstName) + pure(" ")) +
  expToAuthorOps(author).lastName
\end{lstlisting}

\code{Author} is not part of {\LoS} or the standard Scala library, but an application-specific class, hence {\LoS} cannot
pre-define the necessary lifting code. Instead, the application programmer needs
to provide this code to connect {\LoS} to his application. To support the
application programmer with this tedious task, we provide a code generator which
discovers the interface of a class through reflection on its compiled version
and generates the boilerplate code such as the one above for \code{Author}. It
also generates
the application-specific expression tree types such as \code{AuthorFirstName} as shown in Sec.~\ref{subsec:repexptrees}.
In general, query writers need to generate and import the boilerplate lifting code for all application-specific types they want to use in a {\LoS} query.

If desired, we can exclude some methods to \emph{restrict} the language supported in our deeply embedded programs. For instance, \LoS{} requires the user to write side-effect-free queries, hence we do not lift methods which perform side effects.

@@

An expression like \code{"foo" + "bar" + str1} is converted to
\begin{lstlisting}
toStringOps("foo" + "bar") + str1
\end{lstlisting}
Hence, part of the expression is evaluated before being reified. This is harmless here since we want \code{"foo" + "bar"} to be evaluated at compile-time, that is constant-folded, but in other cases it is preferable to prevent the constant folding. We will see later examples where queries on collections are evaluated before reification, defeating the purpose of our framework, and how we work around those.

@@

%\begin{lstlisting}
%def toFOAS[S, T](funBody: Exp[S] => Exp[T]) = 
%\end{lstlisting}
In our example, \code{Fun(str => str + "!")} produces\\
\code{FOASFun(StringConcat(TypedVar[String](1), Const("!")), TypedVar[String](1))}%
%: since an unbound variable appears here, this can be easily recognized as an open term
.
